Reanalysis Step 8: Patient-Level Table 1

### 2025-09-05 reanalysis-8-table1.qmd
better tqables

@@

@@

@@

@@

@@

@@

@@

  # Display totals
  cat("**Total Patients**\n")
  print(kable(table_data$totals))

@@

  cat("\n*Sex*\n")
  print(kable(table_data$sex))

  cat("\n*Age (years)*\n")
  print(kable(table_data$age))

  cat("\n*Race/Ethnicity*\n")
  print(kable(table_data$race))

  cat("\n*Insurance*\n")
  print(kable(table_data$insurance))

@@

  cat("\n*Severity of DR*\n")
  print(kable(table_data$dr))

  cat("\n*DR Treatments (% with treatment)*\n")
  treatments_df <- data.frame(
    Treatment = c("PRP", "Focal laser", "Anti-VEGF", "Other injections"),
    Lapse = c(
      sprintf("%d (%.0f%%)", table_data$treatments$prp$n[1], table_data$treatments$prp$pct[1]),
      sprintf("%d (%.0f%%)", table_data$treatments$focal$n[1], table_data$treatments$focal$pct[1]),
      sprintf("%d (%.0f%%)", table_data$treatments$vegf$n[1], table_data$treatments$vegf$pct[1]),
      sprintf("%d (%.0f%%)", table_data$treatments$other$n[1], table_data$treatments$other$pct[1])
    ),
    NoLapse = c(
      sprintf("%d (%.0f%%)", table_data$treatments$prp$n[2], table_data$treatments$prp$pct[2]),
      sprintf("%d (%.0f%%)", table_data$treatments$focal$n[2], table_data$treatments$focal$pct[2]),
      sprintf("%d (%.0f%%)", table_data$treatments$vegf$n[2], table_data$treatments$vegf$pct[2]),
      sprintf("%d (%.0f%%)", table_data$treatments$other$n[2], table_data$treatments$other$pct[2])
    )
  )
  print(kable(treatments_df, col.names = c("Treatment", "Lapse in care", "No lapse in care")))

  cat("\n*Ophthalmologic Comorbidities*\n")
  comorb_df <- data.frame(
    Condition = c("Glaucoma", "Prior cataract surgery", "Other retinal disease"),
    Lapse = c(
      sprintf("%d (%.0f%%)", table_data$comorbidities$glaucoma$n[1], table_data$comorbidities$glaucoma$pct[1]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$cataract$n[1], table_data$comorbidities$cataract$pct[1]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$retina$n[1], table_data$comorbidities$retina$pct[1])
    ),
    NoLapse = c(
      sprintf("%d (%.0f%%)", table_data$comorbidities$glaucoma$n[2], table_data$comorbidities$glaucoma$pct[2]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$cataract$n[2], table_data$comorbidities$cataract$pct[2]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$retina$n[2], table_data$comorbidities$retina$pct[2])
    )
  )
  print(kable(comorb_df, col.names = c("Condition", "Lapse in care", "No lapse in care")))

  cat("\n**Clinical Characteristics (Mean ± SD)**\n")
  clinical_df <- data.frame(
    Variable = c("Baseline Visual Acuity (logMAR)", 
                "Diabetes Complication Severity Index", 
                "Charlson Comorbidity Index"),
    Lapse = c(table_data$continuous$va$stat[1],
             table_data$continuous$dcsi$stat[1],
             table_data$continuous$cci$stat[1]),
    NoLapse = c(table_data$continuous$va$stat[2],
               table_data$continuous$dcsi$stat[2],
               table_data$continuous$cci$stat[2]),
    Total = c(table_data$continuous$va$stat[3],
             table_data$continuous$dcsi$stat[3],
             table_data$continuous$cci$stat[3])
  )
  print(kable(clinical_df, col.names = c("Variable", "Lapse in care", "No lapse in care", "Total")))
}

format_stratified_table(complete_table_data, "Table 1: Baseline Characteristics by Lapse Status (Complete Cases)")
```

@@

# Display the imputed table
format_stratified_table(imputed_table_data, "Table 2: Baseline Characteristics by Lapse Status (Imputed Data - Averaged)")
```

>>>>

  # Add overall total for treatments
  prp_overall <- data |>
    summarise(
      n = sum(PRP_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      lapse_group = "Total"
    )
  prp_total <- bind_rows(prp_total, prp_overall)

@@

  focal_overall <- data |>
    summarise(
      n = sum(focal_laser_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      lapse_group = "Total"
    )
  focal_total <- bind_rows(focal_total, focal_overall)

@@

  vegf_overall <- data |>
    summarise(
      n = sum(anti_VEGF_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      lapse_group = "Total"
    )
  vegf_total <- bind_rows(vegf_total, vegf_overall)

@@

  other_overall <- data |>
    summarise(
      n = sum(other_inject_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      lapse_group = "Total"
    )
  other_inj_total <- bind_rows(other_inj_total, other_overall)

@@

  glaucoma_overall <- data |>
    summarise(
      n = sum(glaucoma_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      lapse_group = "Total"
    )
  glaucoma_total <- bind_rows(glaucoma_total, glaucoma_overall)

@@

  cataract_overall <- data |>
    summarise(
      n = sum(catsurg_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      lapse_group = "Total"
    )
  cataract_total <- bind_rows(cataract_total, cataract_overall)

@@

  retina_overall <- data |>
    summarise(
      n = sum(otherretina_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      lapse_group = "Total"
    )
  retina_total <- bind_rows(retina_total, retina_overall)

@@

  # Display totals
  cat("**Total Patients**\n")
  knitr::kable(table_data$totals)

@@

  cat("\n*Sex*\n")
  knitr::kable(table_data$sex)

  cat("\n*Age (years)*\n")
  knitr::kable(table_data$age)

  cat("\n*Race/Ethnicity*\n")
  knitr::kable(table_data$race)

  cat("\n*Insurance*\n")
  knitr::kable(table_data$insurance)

@@

  cat("\n*Severity of DR*\n")
  knitr::kable(table_data$dr)

  cat("\n*DR Treatments (% with treatment)*\n")
  treatments_df <- data.frame(
    Treatment = c("PRP", "Focal laser", "Anti-VEGF", "Other injections"),
    Lapse = c(
      sprintf("%d (%.0f%%)", table_data$treatments$prp$n[1], table_data$treatments$prp$pct[1]),
      sprintf("%d (%.0f%%)", table_data$treatments$focal$n[1], table_data$treatments$focal$pct[1]),
      sprintf("%d (%.0f%%)", table_data$treatments$vegf$n[1], table_data$treatments$vegf$pct[1]),
      sprintf("%d (%.0f%%)", table_data$treatments$other$n[1], table_data$treatments$other$pct[1])
    ),
    NoLapse = c(
      sprintf("%d (%.0f%%)", table_data$treatments$prp$n[2], table_data$treatments$prp$pct[2]),
      sprintf("%d (%.0f%%)", table_data$treatments$focal$n[2], table_data$treatments$focal$pct[2]),
      sprintf("%d (%.0f%%)", table_data$treatments$vegf$n[2], table_data$treatments$vegf$pct[2]),
      sprintf("%d (%.0f%%)", table_data$treatments$other$n[2], table_data$treatments$other$pct[2])
    ),
    Total = c(
      sprintf("%d (%.0f%%)", table_data$treatments$prp$n[3], table_data$treatments$prp$pct[3]),
      sprintf("%d (%.0f%%)", table_data$treatments$focal$n[3], table_data$treatments$focal$pct[3]),
      sprintf("%d (%.0f%%)", table_data$treatments$vegf$n[3], table_data$treatments$vegf$pct[3]),
      sprintf("%d (%.0f%%)", table_data$treatments$other$n[3], table_data$treatments$other$pct[3])
    )
  )
  knitr::kable(treatments_df, col.names = c("Treatment", "Lapse in care", "No lapse in care", "Total"))

  cat("\n*Ophthalmologic Comorbidities*\n")
  comorb_df <- data.frame(
    Condition = c("Glaucoma", "Prior cataract surgery", "Other retinal disease"),
    Lapse = c(
      sprintf("%d (%.0f%%)", table_data$comorbidities$glaucoma$n[1], table_data$comorbidities$glaucoma$pct[1]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$cataract$n[1], table_data$comorbidities$cataract$pct[1]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$retina$n[1], table_data$comorbidities$retina$pct[1])
    ),
    NoLapse = c(
      sprintf("%d (%.0f%%)", table_data$comorbidities$glaucoma$n[2], table_data$comorbidities$glaucoma$pct[2]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$cataract$n[2], table_data$comorbidities$cataract$pct[2]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$retina$n[2], table_data$comorbidities$retina$pct[2])
    ),
    Total = c(
      sprintf("%d (%.0f%%)", table_data$comorbidities$glaucoma$n[3], table_data$comorbidities$glaucoma$pct[3]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$cataract$n[3], table_data$comorbidities$cataract$pct[3]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$retina$n[3], table_data$comorbidities$retina$pct[3])
    )
  )
  knitr::kable(comorb_df, col.names = c("Condition", "Lapse in care", "No lapse in care", "Total"))

  cat("\n**Clinical Characteristics (Mean ± SD)**\n")
  clinical_df <- data.frame(
    Variable = c("Baseline Visual Acuity (logMAR)", 
                "Diabetes Complication Severity Index", 
                "Charlson Comorbidity Index"),
    Lapse = c(table_data$continuous$va$stat[1],
             table_data$continuous$dcsi$stat[1],
             table_data$continuous$cci$stat[1]),
    NoLapse = c(table_data$continuous$va$stat[2],
               table_data$continuous$dcsi$stat[2],
               table_data$continuous$cci$stat[2]),
    Total = c(table_data$continuous$va$stat[3],
             table_data$continuous$dcsi$stat[3],
             table_data$continuous$cci$stat[3])
  )
  knitr::kable(clinical_df, col.names = c("Variable", "Lapse in care", "No lapse in care", "Total"))
}

format_stratified_table(complete_table_data, "Table 1: Baseline Characteristics by Lapse Status (Complete Cases)")

# Export complete cases table to CSV
export_stratified_table <- function(table_data, filename) {
  # Initialize empty dataframe for export
  export_df <- data.frame(
    variable = character(),
    lapse_value = numeric(),
    lapse_prop = numeric(),
    lapse_sd = numeric(),
    no_lapse_value = numeric(),
    no_lapse_prop = numeric(),
    no_lapse_sd = numeric(),
    total_value = numeric(),
    total_prop = numeric(),
    total_sd = numeric(),
    stringsAsFactors = FALSE
  )

  # Add total row
  n_lapse <- as.numeric(gsub(" .*", "", table_data$totals$Lapse))
  n_no_lapse <- as.numeric(gsub(" .*", "", table_data$totals$NoLapse))
  n_total <- as.numeric(gsub(" .*", "", table_data$totals$Total))

  export_df <- rbind(export_df, data.frame(
    variable = "Total",
    lapse_value = n_lapse,
    lapse_prop = as.numeric(gsub(".*\\((.*)%.*", "\\1", table_data$totals$Lapse)),
    lapse_sd = NA,
    no_lapse_value = n_no_lapse,
    no_lapse_prop = as.numeric(gsub(".*\\((.*)%.*", "\\1", table_data$totals$NoLapse)),
    no_lapse_sd = NA,
    total_value = n_total,
    total_prop = 100,
    total_sd = NA
  ))

  # Process categorical variables from table_data
  # This is a simplified version - you'll need to expand this based on your exact structure

  # Add treatments
  for (i in 1:3) {  # Lapse, No Lapse, Total
    if (i <= nrow(table_data$treatments$prp)) {
      # Extract values for each treatment
      # This needs more sophisticated parsing based on actual data structure
    }
  }

  # Add continuous variables
  for (var_name in c("baseline_VA_logMAR", "DCSI", "CCI")) {
    var_label <- switch(var_name,
      "baseline_VA_logMAR" = "Baseline VA (logMAR)",
      "DCSI" = "Diabetes Complication Severity Index",
      "CCI" = "Charlson Comorbidity Index"
    )

    # Parse mean ± SD format
    if (var_name == "baseline_VA_logMAR") {
      stats <- table_data$continuous$va$stat
    } else if (var_name == "DCSI") {
      stats <- table_data$continuous$dcsi$stat
    } else {
      stats <- table_data$continuous$cci$stat
    }

    # Parse "mean ± sd" format
    parse_mean_sd <- function(x) {
      parts <- strsplit(x, " ± ")[[1]]
      return(list(mean = as.numeric(parts[1]), sd = as.numeric(parts[2])))
    }

    lapse_stats <- parse_mean_sd(stats[1])
    no_lapse_stats <- parse_mean_sd(stats[2])
    total_stats <- parse_mean_sd(stats[3])

    export_df <- rbind(export_df, data.frame(
      variable = var_label,
      lapse_value = lapse_stats$mean,
      lapse_prop = NA,
      lapse_sd = lapse_stats$sd,
      no_lapse_value = no_lapse_stats$mean,
      no_lapse_prop = NA,
      no_lapse_sd = no_lapse_stats$sd,
      total_value = total_stats$mean,
      total_prop = NA,
      total_sd = total_stats$sd
    ))
  }

  # Save to CSV
  write.csv(export_df, filename, row.names = FALSE)
  cat(paste0("\nTable exported to: ", filename, "\n"))

  return(export_df)
}

# Export the complete cases table
complete_export <- export_stratified_table(complete_table_data, 
                                           file.path(reanalysis_data_dir, "table1_complete_cases.csv"))
```

@@

# Display the imputed table
format_stratified_table(imputed_table_data, "Table 2: Baseline Characteristics by Lapse Status (Imputed Data - Averaged)")

# Export the imputed table
imputed_export <- export_stratified_table(imputed_table_data, 
                                          file.path(reanalysis_data_dir, "table1_imputed.csv"))
```
